---
# Challenge 2

# Name - Maria Master
# INST 314 (0202)
# Dr. Scott Jackson
# GTA: Aditya Patkar
---

```{r}
library(ggplot2) 
```

# Part 1

1. Use a function in R or Python to generate a large number of values from a normal distri-
bution. For this example, let’s use the “standard normal” distribution, which has a mean of
0 and a standard deviation of 1. This is often the default in functions that generate values,
but try to make it explicit in your code.

```{r}
normal_sample <- rnorm(10000, mean = 0, sd = 1)
```

2. Plot a histogram of the random values from #1. If they don’t look “normal enough”, try a larger
number of values.
```{r}
ggplot(data = NULL, aes(x = normal_sample)) + geom_histogram(binwidth = 0.1)
```
3. Generate a large number of values (the same that you settled on for #1) from a uniform distribution between -3 and 3. (This range is chosen to be comparable to the random normal data, for reasons described below.)

```{r}
uniform_sample <- runif(10000, min = -3, max = 3)
```

4. Plot a histogram of the random values from #3. If the distribution doesn’t look “uniform enough”,
try a larger number of values.

```{r}
ggplot(data = NULL, aes(uniform_sample)) + geom_histogram()
```
5. Using random numbers is convenient and useful for a lot of things, but sometimes we also
want to calculate precise values. The histograms above give good approximate shapes for the
distributions, but now we want to be able to plot the exact shape of each distribution. In order
to do that, we need to calculate probability densities for a bunch of specific values of data.

• For example, the probability density from a standard normal distribution for the value of 0
is just under 0.4. In order to understand what these numbers mean, we can think about it
like we have a bag of numbers (our data) and we are trying to understand how likely it is
that we might draw a particular number. It’s a bit more mathematically complex when we
have continuous distributions like the normal distribution, but the intuition is similar. So
when we compute the probability density for a particular data value, like 0, that means we
are asking “how likely is it that I’ll pull a 0 from the data bag?”.

• In order to have enough probability density values to plot a nice distribution curve, we have
to compute the densities for a bunch of data values, across the range of the data. So for
example, if the data generally goes from 1 to 100, we’d want to get the density values for 1,
1.5, 2, 2.5, and so on, up to 100. Or if we wanted a more fine-grained plot, we could go
by smaller increments, like 0.2 or 0.1. This sequence of steps is essentially forming a “grid”
along the x-axis of our plot.

• The range of a normal distribution is infinite, but we obviously can’t create a grid over
infinite values, so we need to pick a reasonable range. For a standard normal distribution
(mean of 0, standard deviation of 1), values rarely go below -4 or above 4. So for this
exercise, we will generate the “grid” of data values by creating the sequence of numbers
from -4 to 4, incrementing by 0.1. These are the values that will be on the x-axis of our
density plot.

```{r}
x_grid <- seq(from = -4, to = 4, by = 0.1)
```
6. After you’ve run the code to create a sequence of data values (from -4 to 4, by 0.1) and calculate the densities for those, you can use those two series of numbers as your x- and y-axis variables, respectively. Create a line plot that uses those values, and you should should have a nice curve that represents the theoretical normal distribution. How does this shape compare to the shape of the histogram you created in #2?

```{r}
ggplot(data = NULL, aes(x = x_grid,
                        y = dnorm(x_grid))) +
                    geom_line() + xlab("x") + ylab("density")
```

*Ans*: Overall, the shape for this probability density plot for the normal distribution and the previous histogram for the random values sample from a standard normal distribution are similar. They are both bell shaped, representing the standard normal distribution. Their symmetry lies around the mean (mean = 0). The values on the tails on both sides as they move away from the mean becomes smaller and smaller. The differences would be that the probability density plot is smoother than the previous histogram. The binning width of the previous histogram does not make plot as smooth as the probability densive curve which is continuous with no binning issue. The scaling is different for both as well. The probability density plot is scaled so that the area under the under sums up to 1, whereas the y-vlalues in the previous histogram presents counts.

7. Using the same process in #5, compute probability densities from a uniform distribution that
goes from -3 to 3, but use the same “grid” (sequence of data values) as above, going from -4 to
4. These ranges are chosen here so that you can visualize the “ends” of the uniform distribution,
since densities will drop abruptly to 0 outside the distribution range. 

```{r}
x_values_uniform <- seq(from = -4, to = 4, by = 0.1)

uniform_densities <- dunif(x_values_uniform, min = -3, max = 3)
```

8. Plot the uniform densities from #7 the same way you plotted the normal densities in 6. Compare
this shape to the shape of your histogram from #4.

```{r}
ggplot(data = NULL, aes(x = x_values_uniform, 
       y = uniform_densities)) +
  geom_line() + xlab("value") + ylab("dentisy")
```
*Ans*: The density plot is smoother than the previous histogram. The uniform distribution is very evident in the second density plot. Outside the range of -3 to 3 in the density curve, the values drop to a 0. This also happens in the histogram but this behavior is clearer in then density curve. Overall, the histogram visualizes the actual distribution of the random values that were generated and it looks very close to a uniform distribution. However, the probability density curve shows the theoretical probability distribution.

9. Now that we’ve seen that histograms of a large number of random numbers are pretty close
to their theoretical distributions, let’s examine the distributions of small numbers of random
numbers. Generate a small number of random numbers from a normal distribution.

```{r}
small_normal_sample <- rnorm(1000, mean = 0, sd = 1)
```

10. Now create a combined plot, that plots both a histogram of the numbers from #9, with a super-
imposed density curve like that from #6. This makes it easier to visually compare the distribution
in your random numbers to the theoretical normal distribution. Hint: you will need to adjust
the histogram so that it is on the density scale. How does the small number of random normal
values compare to the theoretical distribution? Re-run your code a few times, and see how much
it changes each time. 

```{r}
ggplot(data = NULL, aes(x = small_normal_sample)) +
    geom_histogram(aes(y = ..density..), fill = "gray", binwidth = 0.1) +
    geom_density()
```
*Ans*: As the sample size is really small (100) compared to a larger sample size like the previous plots (10,000), the shape is not as bell-shaped as the previous plot. There are gaps between values, high fluctuations and more variability in the sample for the small random sample.

11. Let’s do the same with the uniform distribution, so generate a small number of random numbers
from the same uniform distribution as before (see #3).

```{r}
uniform_sample_small <- runif(100, min = -3, max = 3)
```

12. Now create a similar plot as in #10 with the uniform variables, by plotting a histogram of the
numbers from #11, and superimposing the density curve from #8. How does the distribution from
the small number of random numbers compare to the theoretical uniform distribution? Re-run
the code from #11 and #12 a few times to see how much it changes each time.

```{r}
ggplot(data = NULL, aes(x = uniform_sample_small)) +
    geom_histogram(aes(y = ..density..), fill = "gray", binwidth = 0.1) +
    geom_density()
```
*Ans*: The plot for the small random sample does not highly resemble the uniform distribution for the previous plots with larger sample size. There is more variability in the values and gaps. Some of the numbers are extremely high whereas some are really low. 

# Part 2
13. Let’s start by simulating a simple generating process, like rolling a die. But this time, let’s roll a bigger die! If you’ve ever played particularly nerdy board games, you might know that there are dice with more sides than a standard six-sided cube die. One popular shape is a 20-sided die,
often called a “d20” in gaming parlance.

• Start this section by creating a random sample of many (at least 10,000) numbers drawn
from the possible numbers that could be generated by a 20-sided die.

• Hint: think about whether this is sampling “with replacement” or “without replacement”,
and make sure your code reflects that, or you will likely run into an error

```{r}
many_rolls <- sample(1:20, size = 10000, replace = TRUE)
```

14. Plot a histogram of the numbers from #13. What kind of distribution does this appear to be?

```{r}
#geom_bar() is counting the frequency
ggplot(data = NULL, aes(x = many_rolls)) + geom_bar() +
    scale_x_continuous(breaks = 1:20)
```
*Ans*: The plot seems to have a fairly uniform distribution with each possibility taking place almost the same number of times. 

15. Recall the Central Limit Theorem, which states that when many (independent) processes are
averaged (or summed) together, that average (or sum) tends towards a normal distribution, as
the number of processes increases. How could this apply to our die-rolling simulation? Run an
altered version of your code from #13 to generate numbers that still simulate the rolling of (fair)
20-sided dice, but which result in generating numbers that follow a normal distribution.

```{r}
many_rolls_mean <- mean(many_rolls)
many_rolls_sd <- sd(many_rolls)

many_rolls_normal_data <- rnorm(10000, mean = many_rolls_mean, sd = many_rolls_sd)

many_rolls_normal_data <- many_rolls_normal_data[many_rolls_normal_data >= 1 & many_rolls_normal_data <= 20]
```

16. Now plot a histogram of these new values, to show that they look normal. 

```{r}
ggplot(data = NULL, aes(x = many_rolls_normal_data)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(1, 20, by = 1)) +
  labs(title = "Histogram of Many Rolls' Normal Data",
       x = "many rolls")
```

17. Explain why the change you made in #15 resulted in a normal distribution, despite the fact that
we saw a different distribution in #14. 

*Ans*: The Central Limit Theorem states that "when many (independent) processes are
averaged (or summed) together, that average (or sum) tends towards a normal distribution, as
the number of processes increases." In our case, if we rolls a 20-sided die many many times and sum the outcomes repeatedly, we see a normal distribution. On the other hand, for the previous uniform distribution, we were only counting the distribution of individual die rolls. 

# Part 3

Load data - Cost of living v2.

```{r}
data <- read.csv('/Users/mariagomesmaster/Documents/INST314/Challenge2/cost-of-living_v2.csv')

if (is.data.frame(data)) {
  cat("Data load successful!\n")
} else {
  cat("Unable to load data from the CSV file.\n")
}
```

18. Select a (numeric) variable of interest from your data set, describe it briefly, and plot a histogram of it.

*Ans*: x54 = Average Monthly Net Salary (After Tax) (USD) is a numeric column. As this data set has a range of grocery and daily life items to calculate the cost of living, it is important to learn more about the average month salary across these cities for a better understanding of living standard in the world.  

```{r}
x54_column <- data$x54
summary(x54_column)
```

```{r}
#plot
ggplot(data, aes(x = x54_column)) + geom_histogram()
```
19. In Part 1 of this challenge, we plotted theoretical density curves by computing exact probability densities over the range of our (simulated) data. We will do that again later, but for now, we want to also plot the empirical probability density. This is the plot that looks kind of like a “smoothed” histogram, with a line rather than the binned bars of a histogram. Create a density plot here, and notice how it follows the shape of your histogram from #18.

```{r}
ggplot(data, aes(x=x54_column)) +
  geom_density() + xlab("Value") + ylab("Density")
```
20.  Return to the histogram and alter the bin width from the default, which should make the bars
narrower or wider. Pick a few different values, and settle on the bin width that you think is the
best for this data.

```{r}
ggplot(data, aes(x = x54_column)) + geom_histogram(binwidth = 100)
```
21. Describe how you decided to pick the values of the bin width. For example, how did you pick the
first number you tried, and how did you go from there?

*Ans*: Earlier, I used summary(x54_column) to take a look at the min, max, mean and other summary statistics for the x54 column particularly. I did this to understand if there were any missing values and any massive outlier. Judging from the min which is 18 USD and the max which is 12821 USD, it was clear that there would be an outlier. Therefore, I decided to make my bin width relatively smaller, which is 100, in order to give a bird's eye view to the right-skewed distribution. 

22. Describe what examining different bin widths made you notice about your data, and why you
liked the one you picked the best.

*Ans*: I had initially picked 50 for the bin width. However, making it 50 or any number less than 50 such as a 10 or a 1, makes it extremely difficult to understand the distinction between each bins. I could have picked a larger number than 100 such as 500, however, I did not pick it because even though it gave the general sense of the distribution, I wanted a more specific view of the distribution while not combining too many values in one bin. 

23. Compute the mean of the variable you have been working with.

```{r}
class(data$x54)
```
```{r}
# remove NA
mean_x54 <- mean(data$x54, na.rm = TRUE)

mean_x54
```

24. Compute the standard deviation of the same variable.

```{r}
sd_x54 <- sd(data$x54, na.rm = TRUE)

sd_x54
```

25.  Compute the median of the same variable.

```{r}
median_x54 <- median(data$x54, na.rm = TRUE)

median_x54
```

26. Compute quartiles (0%, 25%, 50%, 75%, and 100% quantiles) of the same variable.

```{r}
quantiles_x54 <- quantile(data$x54, c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)

quantiles_x54
```

27. Compute the following quantiles of the same data: 2.5%, 5%, 25%, 75%, 95%, and 97.5%.

```{r}
quantiles_x54_specific <- quantile(data$x54, c(0.025, 0.05, 0.25, 0.75, 0.95, 0.975), na.rm = TRUE)

quantiles_x54_specific
```

28. Plot a histogram again, but this time additionally plot two vertical lines that mark the mean and median values you computed in #23 and #25 above.

```{r}
ggplot(data, aes(x=x54_column)) +
      geom_histogram(binwidth = 100) +
      geom_vline(aes(xintercept = mean_x54), color = "red")+
      geom_vline(aes(xintercept = median_x54), color = "blue")
```
29. Do the same as in #28, but plot a density curve (like #19) instead of a histogram.

```{r}
ggplot(data, aes(x=x54_column)) +
  geom_density() + ylab("Density") +
  geom_vline(aes(xintercept = mean_x54), color = "red")+
  geom_vline(aes(xintercept = median_x54), color = "blue")
```
30. Examine the plot in #29, and describe the differences between the mean, median, and the mode
of the density plot.

*Ans*: The mean is the average of the values. In this case, the mean is higher than the median, which is the middle value of the sorted values. In the case of a right-skewed distribution, the median is a better measure than the mean because in this case, the median is not affected by the high outliers. The mode represents the highest frequency of a value. In this case, it is the peak of the distribution, and it is lower then both the mean and the median. If this was a normal distribution, the mean would be almost the same as the median and the mean. 

31. With the histogram and density plot, we can roughly tell how normal the distribution looks, but
the best way to compare this distribution to a normal distribution is to plot both at the same
time. This means we need to compute the appropriate probability densities, as we did in Part 1
of this Challenge. Think about what you need: the “grid” of data values for the x-axis, as well
as the parameters of the normal distribution for computing the densities. Then compute the
appropriate probability densities. 

```{r}
# seq of data values
x_values <- seq(min(data$x54, na.rm = TRUE), max(data$x54, na.rm = TRUE), length.out = 1000)

# probability densities for the normal distribution
normal_density <- dnorm(x_values, mean = mean_x54, sd = sd_x54)
```

32. Use the densities from #31 to plot a normal distribution by itself, and confirm that it looks right.

```{r}
density_df <- data.frame(x = x_values, y = normal_density)

ggplot(density_df) + 
      geom_line(aes(x = x_values, y = normal_density)) +
      xlab("Value") + ylab("density")
```

33. Now combine the histogram (as in #19 or #20) with the theoretical density plot from #32. Re-
member to adjust your histogram to be on the density scale. This superimposed plot should
make it even more clear whether your data follows a normal distribution or not. 

```{r}
ggplot(data, aes(x = x54_column)) +
  geom_histogram(aes(y = ..density..), fill = "gray", binwidth = 100) +
  geom_density() +
  labs(x = "Value", y = "Density")
```
34. Finally, select at least one more numeric variable, different from the one you have been working on so far. If your data has enough variables, pick three new ones to look at. For each of these variable, plot a histogram. Which of the variables looks the most like a normal distribution?
Does this make sense for your data? Does this tell you anything or raise any questions in your
mind?

*Ans*: Columns chosen - x33 - Gasoline (1 liter) (USD), x39 - Fitness Club, Monthly Fee for 1 Adult (USD) , x41 - Cinema, International Release, 1 Seat (USD). 

```{r}
# x33
x33_column <- data$x33
summary(x33_column)
```
```{r}
#x33 plot
ggplot(data, aes(x = x33_column)) + geom_histogram()
```
```{r}
#x39
x39_column <- data$x39
summary(x39_column)
```

```{r}
#x39 plot
ggplot(data, aes(x = x39_column)) + geom_histogram()
```
```{r}
#x41
x41_column <- data$x41
summary(x41_column)
```
```{r}
ggplot(data, aes(x = x41_column)) + geom_histogram()
```
Out of all these three numeric, only the gasoline prices seems to have a fairly normal distribution. The movie theater seat price per adult and the gym membership seems to have a right-skewed distribution. This is mostly because of some very high outliers for both the price for gym membership and the cinemas. As this data set contains data from thousands of cities world wide, it is surprising to learn that the gasoline prices are more or less similar across all regions. As for the right-skewed distributions, it could be because of different qualities or levels of service or due to the competition in different markets.  